R for Clinical Study Reports and Submission

### 2021-09-15 tlf-population.Rmd
update tlf-population

@@

The first step is to read relevant datasets into R. 
For analysis population table, 
all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

@@

```{r}
adsl %>%
  select(USUBJID, ITTFL, EFFFL, SAFFL) %>%
  head(4)
```
## Helper Functions

Before we write analysis code, let's discuss the possibility to
reuse R code by writing some toy helper function.

As discussed in 
[R for data science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
"You should consider writing a function 
whenever you’ve copied and pasted a block of code more than twice".

@@

- format percentages using `formatC` 
- calculate numbers and percentages by groups

We create two ad-hoc functions and apply them to all rest of tables.

To format numbers and percentages, we created a function called `fmt_num`.
It is a very simple function wrapping `formatC`.

@@

The main reason to create `fmt_num` is to enhance readability of the analysis code.

For example, we can compare the two versions of code to format percentage 
used in \@ref(disposition) and `fmt_num`.

@@

To calculate numbers and percentages of patients by groups, 
we provide a simple (but not robust) wrapper function, `count_by` using `dplyr` and `tidyr`.

The function can be enhanced in multiple ways, but here we only focus on simplicity and readability.
More details about writing R functions can be found in [Stat454 course](https://stat545.com/functions-part1.html).

@@

By using `count_by`, we can simplify the analysis code as below.

@@

With the helper functions `count_by`, we can easily prepare report dataset as

@@

Now we combine individual rows into the whole table for reporting purpose. 
`tbl_pop` is used as input for `r2rtf` to create final report.

@@

In conclusion, the procedure to generate a population summary table as shown in the above example is listed as follows:

- Step 1: Read data into R, i.e.,  `adsl`.
- Step 2: Rowly bind the counts/percentages of the ITT population, the efficacy population and the safety population. These these types of population are generated by a help function `count_by`. 
- Step 3: Format the output in Step 2 by `r2rtf`.

>>>>

---
editor_options: 
  markdown: 
    wrap: 72
---

@@

The first step is to read relevant datasets into R. For the analysis
population table, all the required information is saved in the ADSL
dataset. We can use the `haven` package to read the dataset.

@@

```{r}
adsl %>%
  select(USUBJID, ITTFL, EFFFL, SAFFL) %>%
  head(4)
```

## Helper Functions

Before we write analysis code, let's discuss the possibility to reuse R
code by writing some toy helper functions.

As discussed in [R for data
science](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function),
"You should consider writing a function whenever you've copied and
pasted a block of code more than twice".

@@

-   Format the percentages using the `formatC()` function.
-   Calculate the numbers and percentages by groups.

We create two ad-hoc functions and apply them to all the rest of the
tables.

To format numbers and percentages, we created a function called
`fmt_num()`. It is a very simple function wrapping `formatC()`.

@@

The main reason to create the `fmt_num()` function is to enhance the
readability of the analysis code.

For example, we can compare the two versions of code to format the
percentage used in \@ref(disposition) and `fmt_num`.

@@

To calculate the numbers and percentages of patients by groups, we
provide a simple (but not robust) wrapper function, `count_by()` using
the `dplyr` and `tidyr` package.

The function can be enhanced in multiple ways, but here we only focus on
simplicity and readability. More details about writing R functions can
be found in the [Stat454
course](https://stat545.com/functions-part1.html).

@@

By using the `count_by()` function, we can simplify the analysis code as
below.

@@

With the helper functions `count_by`, we can easily prepare a report
dataset as

@@

Now we combine individual rows into the whole table for reporting
purposes. `tbl_pop` is used as input for `r2rtf` to create the final
report.

@@

The procedure to generate an analysis population table can be summarized
as follow:

-   Step 1: Read data into R, i.e., `adsl`.
-   Step 2: Bind the counts/percentages of the ITT population, the
    efficacy population and the safety population by row using the
    `count_by()` function.
-   Step 3: Format the output in Step 2 by `r2rtf`.
